Approximately Exact Calculations for Linear Mixed Models

### 2015-06-02 ms.tex
Andrew's minor edits

\noindent\textbf{Claim} \emph{For any positive number $M$, there exist positive numbers $\widetilde{\sigma_e^2} > \sigesq{}^M$ and $\widetilde{\sigma_s^2} > \sigssq{}^M$ that determine a box $B \supset B_1$ whose lower-left and upper-right corners are $(0,0)$ and $(\widetilde{\sigma_e^2}, \widetilde{\sigma_s^2})$, respectively, such that
\begin{equation*}
  \sup_{B_1^c} \logRLss \le \logRL(\sshat) - M.
\end{equation*}}
I.e., \textbf{D1} is satisfied.\\
\noindent\textbf{Proof}
Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.  The proof considers several regions outside $B_1$ in turn, as exemplified by the points $q_1$, $q_2$, $q_3$, and $q_4$ in Figure~\ref{fig:boundingbox2}.  Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.  Let $q_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point satisfying $\sigesq{}_1 \le \sigma_e^{2*}$ and $\sigssq{}_1 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line.  The partial derivatives (\ref{eq:drlds}), between $q_1$ and $\widetilde{q_1}$ are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but zero for $j=s_z+1$.  The partial derivatives (\ref{eq:drlde}), betweem $q_1$ and $\widetilde{q_2}$ are negative for $j \in \{1, \dots, s_z, s_z+1\}$ but nonnegative for $j=s_z+2$.  Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.  
\begin{equation}
	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
\end{equation}
which is larger than $M$ if
\begin{equation}
\label{eq:q1}
  \begin{split}
    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
  \end{split}
\end{equation}
Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.  Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.

@@

Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.  Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis; let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$; let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$; let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$; and let $\logRL_j$ denote the $j$'th term of \eqref{eq:sum}.  The proof considers in turn four regions $Q_1$, $Q_2$, $Q_3$, and $Q_4$ whose union is $B^c$ and shows that $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied on each region.  

@@

	\logRL(\sshat) - \logRL(q) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),

@@

@@

  $M$ iff $\logRL(\widetilde{q}) < L-M$.  But for any fixed $\sigssq$,

@@

      \frac{\partial\logRL_j}{\partial\sigssq} < 0

@@

For a Bayesian analysis there are 44 lines, as shown in Figure~\ref{fig:hmoBayeslines}.  Figure~\ref{fig:hmoBayeslines} differs from Figure~\ref{fig:hmolines} in that it includes a horizontal line for $j = s_z+2$ and the position of the vertical line for $j = s_z+1$ is slightly shifted.
\begin{figure}
	\centering
	\includegraphics[width=.45\linewidth]{figs/hmolines_HH11_Bayes.pdf}
	\caption{The 44 lines, $j=1, 2, \dots, 44$, for the Bayesian analysis of the HMO data.
	              Lines 43 and 44 are dashed.}
	\label{fig:hmoBayeslines}
\end{figure}
Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$, again with control parameters
\begin{equation*}
	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7.
\end{equation*}
The algorithm finished in 22 iterations and took a little under 3 hours.  Table~\ref{table:hmo_HH11Bayes} shows the output and Figure~\ref{fig:hmoBayes} shows the boxes.
\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-\infty$\\
2 & 16 & 0 & -1313.42\\
3 & 52 & 3 & -1285.05\\
4 & 132 & 22 & -1270.76\\
5 & 264 & 88 & -1263.98\\
6 & 628 & 195 & -1260.47\\
7 & 1780 & 378 & -1258.71\\
8 & 4188 & 1111 & -1257.70\\
9 & 5676 & 3880 & -1256.88\\
10 & 5136 & 8272 & -1255.56\\
11 & 5528 & 12026 & -1254.20\\
12 & 8236 & 15495 & -1252.84\\
13 & 12488 & 20609 & -1251.50\\
14 & 23852 & 27134 & -1250.26\\
15 & 22740 & 45301 & -1249.22\\
16 & 43228 & 57234 & -1248.77\\
17 & 160208 & 60410 & -1248.77\\
18 & 331900 & 137643 & -1248.77\\
19 & 575196 & 325744 & -1248.77\\
20 & 942860 & 665225 & -1248.77\\
21 & 722412 & 1427482 & -1248.77\\
22 & 0 & 2149894 & -1248.77\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 22 iterations for the Bayesian analysis of the HMO data.}
\label{table:hmo_HH11Bayes}
\end{table}
The Bayesian analysis in Figure~\ref{fig:hmoBayes} can be compared to the $\logRLorig$ analysis in Figure~\ref{fig:hmo1}.  The peak of the log posterior is near $(600, .05)$, very far from the peak of the $\logRLorig$ around $(500,100)$.  The posterior peak is due to the $\IG(1.1,0.1)$ prior for $\sigssq$, which has a mean of 1, an infinite variance, and a peak at 0.048.  The region near $(500,100)$, though having a lower posterior density, is part of a nearly flat plateau covering a large area.  The MCMC draws depicted in \cite{henn&hodges:2014} Figure~2c show that the plateau has significant posterior mass.
%Usual analyses of the posterior density are carried out either by MCMC, in which case one hopes that one's Markov chain happens to sample all regions of high enough posterior density to matter, or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough.  With our analysis there is no need to hope; the algorithm is certain to find all regions of high density.

@@

%\begin{figure}
%  \begin{subfigure}{.5\textwidth}
%    \centering
%    \includegraphics[width=.8\linewidth]{figs/gmst-boxes2.jpg}
%    \caption{Locations of boxes}
%    \label{fig:gmst-boxes2}
%  \end{subfigure}
%  \begin{subfigure}{.5\textwidth}
%    \centering
%    \includegraphics[width=.8\linewidth]{figs/gmst-rll2.jpg}
%    \caption{Grayscale shows $L^b$ in each box.
%                  Red dot shows $b^* \equiv\argmax L^b$.
%	          Boxes outlined in blue have $U^b \ge L^{b^*}$.
%	          Axes are logarithmic.}
%    \label{fig:gmstrll2}
%  \end{subfigure}
%  \caption{The 3,605,095 boxes from the second analysis of global
%                mean surface temperature.}
%\label{fig:gmst2}
%\end{figure}

\section{Discussion}
This paper explains and illustrates an algorithm for linear mixed models with two variances, to find all regions where either the restricted likelihood function or the joint posterior density of the variances is high, and to evaluate the function there to arbitrary accuracy.  A natural question to ask is \emph{What about linear mixed models with more than two variances?}  A partial answer is given by \cite{hodges:2013} who shows that some models with more than two variances can be re-expressed similarly to \eqref{eq:reexpress} but others can't.  More complex models that can be re-expressed this way include, but are probably not limited to, models displaying general balance that are also orthogonal designs \citep[all balanced ANOVAs plus other models;][]{houtman_speed:1983}, models that are separable in a specific sense \citep[Section 17.1.5]{hodges:2013}, and miscellaneous other models \citep[Section 17.1.5]{hodges:2013}, e.g., a spatial model including random effects for heterogeneity and spatial clustering (an improper conditional autoregressive effect). We have not explored whether the re-expressible models can be analyzed by our algorithm; that's one direction for future work.

>>>>

\noindent\textbf{Claim} \emph{For any positive number $M$, there exist positive numbers $\widetilde{\sigma_e^2} > \sigesq{}^M$ and $\widetilde{\sigma_s^2} > \sigssq{}^M$ that determine a box $B \supset B_1$ whose lower-left and upper-right corners are $(0,0)$ and $(\widetilde{\sigma_e^2}, \widetilde{\sigma_s^2})$, respectively, such that
\begin{equation*}
  \sup_{B^c} \logRLss \le \logRL(\sshat) - M.
\end{equation*}}
I.e., \textbf{D1} is satisfied.\\
\noindent\textbf{Proof}
Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.  The proof considers several regions outside $B$ in turn, as exemplified by the points $q_1$, $q_2$, $q_3$, and $q_4$ in Figure~\ref{fig:boundingbox2}.  Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.  Let $q_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point satisfying $\sigesq{}_1 \le \sigma_e^{2*}$ and $\sigssq{}_1 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line.  The partial derivatives (\ref{eq:drlds}), between $q_1$ and $\widetilde{q_1}$ are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but zero for $j=s_z+1$.  The partial derivatives (\ref{eq:drlde}), betweem $q_1$ and $\widetilde{q_2}$ are negative for $j \in \{1, \dots, s_z, s_z+1\}$ but nonnegative for $j=s_z+2$.  Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.  
\begin{equation}
	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
\end{equation}
which is larger than $M$ if
\begin{equation}
\label{eq:q1}
  \begin{split}
    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
  \end{split}
\end{equation}
Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.  Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.

@@

Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.  Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$, the intercept of $B$ and the $\sigssq$ axis; let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$; let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$; let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$, the intercept of $B$ and the $\sigssq$ axis; and let $\logRL_j$ denote the $j$'th term of \eqref{eq:sum}.  The proof considers in turn four regions $Q_1$, $Q_2$, $Q_3$, and $Q_4$ whose union is $B^c$.  Implicitly, $B$ and $B^c$ are functions of $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$.  The proof shows that $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied on each region.  

@@

	\logRL(\sshat) - \logRL(q) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}).

@@

  The r.h.s.~is larger than $M$ if

@@

  $M$ iff $\logRL(\widetilde{q_3}) < L-M$.  But for any fixed $\sigssq$,

@@

      \frac{\partial\logRL_j}{\partial\sigesq} < 0 \hspace{1cm} \text{and} \hspace{1cm}

@@

For a Bayesian analysis there are 44 lines, as shown in Figure~\ref{fig:hmoBayeslines}.  Figure~\ref{fig:hmoBayeslines} differs from Figure~\ref{fig:hmolines} in that it includes a horizontal line for $j = s_z+2$ and the position of the vertical line for $j = s_z+1$ is slightly shifted.
\begin{figure}
	\centering
	\includegraphics[width=.45\linewidth]{figs/hmolines_HH11_Bayes.pdf}
	\caption{The 44 lines, $j=1, 2, \dots, 44$, for the Bayesian analysis of the HMO data.
	              Lines 43 and 44 are dashed.}
	\label{fig:hmoBayeslines}
\end{figure}
Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$, again with control parameters
\begin{equation*}
	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7.
\end{equation*}
The algorithm finished in 22 iterations and took a little under 3 hours.  Table~\ref{table:hmo_HH11Bayes} shows the output and Figure~\ref{fig:hmoBayes} shows the boxes.
\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-\infty$\\
2 & 16 & 0 & -1313.42\\
3 & 52 & 3 & -1285.05\\
4 & 132 & 22 & -1270.76\\
5 & 264 & 88 & -1263.98\\
6 & 628 & 195 & -1260.47\\
7 & 1780 & 378 & -1258.71\\
8 & 4188 & 1111 & -1257.70\\
9 & 5676 & 3880 & -1256.88\\
10 & 5136 & 8272 & -1255.56\\
11 & 5528 & 12026 & -1254.20\\
12 & 8236 & 15495 & -1252.84\\
13 & 12488 & 20609 & -1251.50\\
14 & 23852 & 27134 & -1250.26\\
15 & 22740 & 45301 & -1249.22\\
16 & 43228 & 57234 & -1248.77\\
17 & 160208 & 60410 & -1248.77\\
18 & 331900 & 137643 & -1248.77\\
19 & 575196 & 325744 & -1248.77\\
20 & 942860 & 665225 & -1248.77\\
21 & 722412 & 1427482 & -1248.77\\
22 & 0 & 2149894 & -1248.77\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 22 iterations for the Bayesian analysis of the HMO data.}
\label{table:hmo_HH11Bayes}
\end{table}
The Bayesian analysis in Figure~\ref{fig:hmoBayes} can be compared to the $\logRLorig$ analysis in Figure~\ref{fig:hmo1}.  The peak of the log posterior is near $(600, .05)$, very far from the peak of the $\logRLorig$ around $(500,100)$.  The posterior peak is due to the $\IG(1.1,0.1)$ prior for $\sigssq$, which has a mean of 1, an infinite variance, and a peak at 0.048.  The region near $(500,100)$, though having a lower posterior density, is part of a nearly flat plateau covering a large area.  The MCMC draws depicted in \cite{henn&hodges:2014} Figure~2c show that the plateau has significant posterior mass.

@@

\section{Discussion}
This paper explains and illustrates an algorithm that facilitates computation for linear mixed models with two variances. The algorithm finds all regions where either the restricted likelihood function or the joint posterior density of the variances is high and can evaluate the function there to arbitrary accuracy.  A natural question to ask is \emph{What about linear mixed models with more than two variances?}  A partial answer is given by \cite{hodges:2013} who shows that some models with more than two variances can be re-expressed similarly to \eqref{eq:reexpress} but others can't.  More complex models that can be re-expressed this way include, but are probably not limited to, models displaying general balance that are also orthogonal designs \citep[all balanced ANOVAs plus other models;][]{houtman_speed:1983}, models that are separable in a specific sense \citep[Section 17.1.5]{hodges:2013}, and miscellaneous other models \citep[Section 17.1.5]{hodges:2013}, e.g., a spatial model including random effects for heterogeneity and spatial clustering (an improper conditional autoregressive effect). We have not explored whether the re-expressible models can be analyzed by our algorithm; that's one direction for future work.
